Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2007-07-08 SCP2008/paper/metamonads.tex
Fix parts of the bibliography (and adjust some references). Minor edits.  Add some comments on things that still need to be fixed.

A survey \cite{carette04} of
Gaussian elimination implementations in the industrial package Maple
found 6 clearly identifiable aspects and 35 different implementations of the
algorithm, as well as 45 implementations of directly related algorithms such as
LU decomposition, Cholesky decomposition, and so on.  We could
manually write each of these implementations optimizing for particular aspects
and using cut-and-paste to ``share'' similar pieces of code.
Or we can write a very generic procedure that accounts for
all the aspects with appropriate abstractions \cite{Axiom}. The
abstraction mechanisms however -- be they procedure, method or a
function call -- have a significant cost, especially for
high-performance numerical computing \cite{carette04}.

A more appealing approach is generative programming
\cite{Czarnecki,Veldhuizen:1998:ISCOPE,musser94algorithmoriented,BOOST,POOMA,ATLAS}.
The approach is not without problems, e.g., making sure that the
generated code is well-formed. This is a challenge in string-based
generation systems, which generally do not offer any guarantees and
therefore make it very difficult to determine which part of the
generator is at fault when the generated code cannot be parsed. Other
problems is preventing accidental variable capture (so-called hygiene
\cite{HygienicMacros}) and ensuring the generated code is
well-typed. Lisp-style macros, Scheme hygienic macros, the camlp4
preprocessor \cite{camlp4}, C++ template meta-programming, and Template
Haskell \cite{conf/dagstuhl/CzarneckiOST03} solve some of the above
problems. Of the widely available maintainable languages, only
MetaOCaml \cite{CTHL03,metaocaml-org}  solves all the above problems
including the well-typing of both the generator and 
the generated code \cite{TahaSheard97,TahaThesis}.

@@

In this paper, we report on our continued progress \cite{CaretteKiselyov05}
\oleg{should we say in the footnote: ``compared to the earlier,
  conference version of this paper we describe a different
  version of our generator dealing with the complete LU decomposition
  and solving. We worked out previously missing aspects of in-place
  updates, representaing permutation matrix, augmented input matrix,
  and back-propagation. We have changed the representation of
  domain-specific knowledge about permissible compositions of aspects.''}
the code generating LU decomposition and solving 
solving these problems using linear solving and LU decomposition
as our running example. Specifically, our contributions:
\begin{itemize}
    \item Extending a let-insertion, memoizing monad of
      \cite{MSP:PADL04,KiselyovTaha} for generating control structures
      such as loops and conditionals. The extension is non-trivial
      because of control dependencies and because
      let-insertion, as we argue, is a control effect on its own.
      : e.g.,\\
      |let x = exp in ...| has a different \emph{effect} within a
      conditional branch.
    \item Implementation of the |perform|-notation (patterned after
      |do|-notation of Haskell) to make monadic code readable.
    \item Use of functors (including higher-order functors) to
      modularize the generator, express aspects (including results of
      various types) and \emph{assure composability of aspects} even
      for aspects that use state and have to be accounted in many
      places in the generated code.
    \item Encode domain-specific knowledge in the generators which 
      will catch domain-specific instantiation errors at generation
      time.
    \item Provide a thorough breakdown of the family of LU decomposition
      algorithms.
\end{itemize}

@@

Before studying the details of the technologies involved in the 
implementation, it is worthwhile to carefully study the design space involved.
A preliminary study~\cite{carette04} revealed a number of aspects of the 
family of Gaussian Elimination algorithms.  In the present work, we outline
a number of additional aspects involved in the faimily of LU decomposition 
algorithms.  These will first be presented in a somewhat ad hoc manner, 
roughly coresponding to the order in which they were ``discovered''.  These
will then be reorganized into sets of semantically related aspects, which will
form the basis of our design.  Finally, we will extract a methodology from
our experience.

@@

We reuse the english word ``aspect'' for the various facets of the family
of LU decomposition algorithms.  While this use if very much in the same 
spirit as in aspect-oriented programming (AOP)~\cite{Kiczales}, our 
implementation methodology is radically different.  We firmly believe that
our typed generative methodogy is much better suited to the functional
programming paradigm than attempting to graft the program-trace-based
methodology of object-oriented versions of AOP.

At this point in time, it is better to think of aspects as purely
design-time entities.  Here we are firmly influenced by Parnas' original
view of modules and information hiding~\cite{journals/cacm/parnas72a} as well
as his view of product families \cite{Parnas??}, and by Dijkstra's ideas on
separation of concerns \cite{EWD:EWD447}.
To apply these principles to the study of LU decomposition, we need
to understand what are the changes between different implementations, and 
what concerns need to be addressed.  We also need to study the degree
to which these concerns are independent.

@@

		one chooses a $PLU$, $PLU^{1}R$ or \emph{Cholesky} 

@@

		See~\cite{ZhCaJeMo07} for an example.

@@

We use MetaOCaml brackets |.<...>.| to generate code expressions,
i.e., to construct future-stage computations. MetaOCaml provides only
one mechanism of combining code expressions, by inlining
\oleg{splicing?} one into
another. The power of that operation, called escape |.~|, comes from
the fact that the expression to be spliced in (inlined) can be
computed: escape lets us perform an arbitrary immediate code-generating
computation \emph{while} we are
building the future-stage computation. The immediate computation in
|simplest_code| is the evaluation of the function |gen|, which in turn
applies |plus|. The function |gen| receives code expressions |.<x>.|
and |.<y>.| as arguments. At the generating stage, we can manipulate
code expressions as (opaque) values. The function |gen| returns a code
expression, which is inlined in the place of the escape. MetaOCaml can
print out code expressions, so we can see the final generated code. It
has no traces of |gen| and |plus|: their applications are done at the
generation stage.

@@

The key is to use continuation-passing style (CPS). Its benefits were
first pointed out by \cite{Bondorf:92} in the context of partial
evaluation, and extensively used by \cite{MSP:PADL04,KiselyovTaha} for
code generation. Now, |param_code2 plus one| gives us the desired
code.

@@

\subsection{Domains}
Reviewing the various aspects outlined in section~\ref{aspects}, the
simplest parametrization is to make the domain abstract.  As it turns
out, we need the following to exist in our domains: $0$, $1$, $+$,
$*$, (unary and binary) $-$, at least \emph{exact} division,
normalization, and potentially a relative size measure. The simplest
case of such domain abstraction is |param_code1|.  There,
code-generators such as |plus| and |one| were passed as arguments. We
need far more than two parameters, so we have to group them. Instead
of the grouping offered by regular records, we use OCaml
\emph{structures} (i.e., modules) so we can take advantage of
extensibility, type abstraction and constraints, and especially
parameterized structures (\emph{functors}).  We define the type of the
domain, the signature |DOMAIN|, which different domains must satisfy:

@@

We chose therefore a different way of maintaining the generator state,
using a monad. We need monad anyway for let-insertion; we add state as
an additional monad parameter (Fig.~\ref{our-monad}) and monadic
actions |fetch| and |store| to access that monadic state. The state is
threaded throughout the entire code-generation computation.

@@

More interesting is the aspect of what to return
from the LU algorithm.  One could create an algebraic data type (as
was done in \cite{carette04}) to encode the various choices: the
matrix, the matrix and the rank, the matrix and the determinant, the
matrix, rank and determinant, and so on. This is wholly unsatisfying
as we know that for any single use, only one of the choices is ever
possible, yet any routine which calls the generated code must deal
with these unreachable options.  Instead we use a module type with an
\emph{abstract} type |res| for the result type; different instances of
the signature set the result type differently. Given below is this
module type and one instantiation, which specifies the output of a LU
algorithm as a 3-tuple |contr * Det.outdet * int| of the U-factor, the
determinant and the rank.

@@

\section{Runge-Kutta solvers}
\label{s:ode}
\oleg{Drop this? no space}

>>>>

A survey \cite{Carette06} of
Gaussian elimination implementations in the industrial package Maple
found 6 clearly identifiable aspects and 35 different implementations of the
algorithm, as well as 45 implementations of directly related algorithms such as
LU decomposition, Cholesky decomposition, and so on.  We could
manually write each of these implementations optimizing for particular aspects
and using cut-and-paste to ``share'' similar pieces of code.
Or we can write a very generic procedure that accounts for
all the aspects with appropriate abstractions \cite{Axiom}. The
abstraction mechanisms however -- be they procedure, method or a
function call -- have a significant cost, especially for
high-performance numerical computing \cite{Carette06}.

A more appealing approach is generative programming
\cite{Czarnecki,Veldhuizen:1998:ISCOPE,musser89generic,musser94algorithmoriented,BOOST,POOMA,ATLAS}.
The approach is not without problems, e.g., making sure that the
generated code is well-formed. This is a challenge in string-based
generation systems, which generally do not offer any guarantees and
therefore make it very difficult to determine which part of the
generator is at fault when the generated code cannot be parsed. Other
problems is preventing accidental variable capture (so-called hygiene
\cite{HygienicMacros}) and ensuring the generated code is
well-typed. Lisp-style macros, Scheme hygienic macros, the camlp4
preprocessor \cite{camlp4}, C++ template meta-programming, and Template
Haskell \cite{conf/dagstuhl/CzarneckiOST03} solve some of the above
problems. Of the widely available maintainable languages, only
MetaOCaml \cite{CTHL03,metaocaml-org}  solves all the above problems
including the well-typing of both the generator and 
the generated code \cite{TahaSheard97,TahaThesis}.

@@

In this paper, we report on our continued progress \cite{CaretteKiselyov05}
\footnote{We describe here an improved version of our generator
  dealing with the complete LU decomposition algorithm, as well as
  linear solving. We worked out previously missing aspects of in-place
  updates, representing permutation matrices, dealing with augmented 
  input matrix, and back-propagation. We have also changed the representation
  of domain-specific knowledge about permissible compositions of aspects.
  Also included is a careful description of all the aspects involved, as
  well as documenting our development methodology for highly parametric
  scientific software.}
in using code generation for scientific (both numeric and symbolic)
software.  We will use the algorithm family of LU decomposition 
and linear system solving as our running examples to demonstrate our
techniques.  Specifically, our contributions:
\begin{itemize}
    \item Extending a let-insertion, memoizing monad of
      \cite{MSP:PADL04,KiselyovTaha} for generating control structures
      such as loops and conditionals. The extension is non-trivial
      because of control dependencies and because
      let-insertion, as we argue, is a control effect on its own.
      : e.g.,\\
      |let x = exp in ...| has a different \emph{effect} within a
      conditional branch.
    \item Implementation of the |perform|-notation (patterned after
      the |do|-notation of Haskell) to make monadic code readable.
    \item Use of functors (including higher-order functors) to
      modularize the generator, express aspects (including results of
      various types) and \emph{insure composability of aspects} even
      for aspects that use state and have to be accounted for in many
      places in the generated code.
    \item Encode domain-specific knowledge in the generators which 
      will catch domain-specific instantiation errors at generation
      time.
    \item Provide a thorough breakdown of the family of LU decomposition
      algorithms.
\end{itemize}

@@

Before studying the details of the technologies involved in the 
implementation, it is worthwhile to carefully study the design space involved.
A preliminary study~\cite{Carette06} revealed a number of aspects of the 
family of Gaussian Elimination algorithms.  In the present work, we outline
a number of additional aspects involved in the faimily of LU decomposition 
algorithms.  These will first be presented in a somewhat ad hoc manner, 
roughly coresponding to the order in which they were ``discovered''.  These
will then be reorganized into sets of semantically related aspects, which will
form the basis of our design.  Finally, we will extract a methodology from
our experience.

@@

We reuse the english word ``aspect'' for the various facets of the family
of LU decomposition algorithms.  While this use if very much in the same 
spirit as in aspect-oriented programming (AOP)~\cite{kiczales97aspectoriented},
our implementation methodology is radically different\footnote{However it seems
that we are closer to the original ideas on AOP~\cite{709568,mendhekar97rg}
which were also concerned with scientific software}.  We firmly believe that
our typed generative methodogy is much better suited to the functional
programming paradigm than attempting to graft the program-trace-based
methodology of object-oriented versions of AOP.

At this point in time, it is better to think of aspects as purely
design-time entities.  Here we are firmly influenced by Parnas' original
view of modules and information hiding~\cite{journals/cacm/parnas72a} as well
as his view of product families \cite{journals/tse/Parnas76}, and by
Dijkstra's ideas on separation of concerns \cite{EWD:EWD447}.
To apply these principles to the study of LU decomposition, we need
to understand what are the changes between different implementations, and 
what concerns need to be addressed.  We also need to study the degree
to which these concerns are independent.

@@

		one chooses a $PLU$, $PLUR$ or \emph{Cholesky} 

@@

		See~\cite{ZhCaJeMo06a} for an example.

@@

We use MetaOCaml brackets |.<...>.| to generate code expressions,
i.e., to construct future-stage computations. MetaOCaml provides only
one mechanism of combining code expressions, by splicing one
piece of code into
another. The power of that operation, called escape |.~|, comes from
the fact that the expression to be spliced in (inlined) can be
computed: escape lets us perform an arbitrary immediate code-generating
computation \emph{while} we are
building the future-stage computation. The immediate computation in
|simplest_code| is the evaluation of the function |gen|, which in turn
applies |plus|. The function |gen| receives code expressions |.<x>.|
and |.<y>.| as arguments. At the generating stage, we can manipulate
code expressions as (opaque) values. The function |gen| returns a code
expression, which is inlined in the place of the escape. MetaOCaml can
print out code expressions, so we can see the final generated code. It
has no traces of |gen| and |plus|: their applications are done at the
generation stage.

@@

The key is to use continuation-passing style (CPS). Its benefits were
first pointed out by \cite{Bondorf:92} in the context of partial
evaluation, and extensively used by \cite{MSP:PADL04,KiselyovTaha} for
code generation. Like \cite{conf/pepm/BondorfD94}, we use this in the 
context of writing a cogen by hand.  Now, |param_code2 plus one| gives us the
desired code.

@@

\subsection{Domains}
Reviewing the various aspects outlined in section~\ref{design}, the
simplest parametrization is to make the domain abstract.  As it turns
out, we need the following to exist in our domains: $0$, $1$, $+$,
$*$, (unary and binary) $-$, at least \emph{exact} division,
normalization, and potentially a relative size measure. The simplest
case of such domain abstraction is |param_code1|.  There,
code-generators such as |plus| and |one| were passed as arguments. We
need far more than two parameters, so we have to group them. Instead
of the grouping offered by regular records, we use OCaml
\emph{structures} (i.e., modules) so we can take advantage of
extensibility, type abstraction and constraints, and especially
parameterized structures (\emph{functors}).  We define the type of the
domain, the signature |DOMAIN|, which different domains must satisfy:

@@

We chose therefore a different way of maintaining the generator state,
using a monad. We need monad anyway for let-insertion; we add state as
an additional monad parameter (Fig.~\ref{ourmonad}) and monadic
actions |fetch| and |store| to access that monadic state. The state is
threaded throughout the entire code-generation computation.

@@

More interesting is the aspect of what to return
from the LU algorithm.  One could create an algebraic data type (as
was done in \cite{Carette06}) to encode the various choices: the
matrix, the matrix and the rank, the matrix and the determinant, the
matrix, rank and determinant, and so on. This is wholly unsatisfying
as we know that for any single use, only one of the choices is ever
possible, yet any routine which calls the generated code must deal
with these unreachable options.  Instead we use a module type with an
\emph{abstract} type |res| for the result type; different instances of
the signature set the result type differently. Given below is this
module type and one instantiation, which specifies the output of a LU
algorithm as a 3-tuple |contr * Det.outdet * int| of the U-factor, the
determinant and the rank.

@@

\section{Runge-Kutta solvers}
\label{s:ode}
\oleg{Drop this? no space}
\jacques{We should drop this as a whole section, but I still think we 
should mention it as something that we have done.  Mostly to say that
things ``extend'' to other families too, and that no new difficulties 
were encountered.}
